Simulations of Ballot Polling Risk-Limiting Audits

### 2023-02-19 providence/sections/pilot.tex
Update pilot.tex

A first round size of $140$ ballots with large probability of stopping ($0.95$) was selected.
Selection order was tracked for the sake of analysis.
As expected, the audit concluded in the first round. The \Providence risk was $0.0418$. Table~\ref{tab:pilot-risks} shows risk measures for the drawn sample using \Providence, \Minerva and \BRAVO (both EoR and SO).

\begin{table}[h!]
\begin{center}
\begin{tabular}{ |c|c|c|c|c| } 
\hline
%\diagbox[dir=NW]{First \\Round \\Size}{RLA}
ballots& \rotatebox{45}{\Providence} & \rotatebox{45}{\Minerva} & \rotatebox{45}{SO \BRAVO} & \rotatebox{45}{EoR \BRAVO} \\
\hline
140 & \bf{0.0418} & \bf{0.0418} & \bf{0.0541} & 0.366 \\
\hline
\end{tabular}
\end{center}
\caption{Risk measures for the drawn first round of $140$ ballots in the Providence, RI pilot audit. Risks in bold meet the risk-limit ($10\%$) and thus correspond to audits that would stop.}
\label{tab:pilot-risks}
\end{table}

@@

For the sake of analysis, the selection order of the ballots sampled during the pilot was also recorded. Figure~\ref{fig:pilot_sequence} shows the cumulative tally of winner ballots after each new ballot in the selection order is added to the sample. We observe two interesting phenomena in this particular sample's selection order.

First, an SO \BRAVO audit of this sample stops because the \BRAVO condition is met when the sample (orange line) surpasses the minimum number of winner ballots (blue line) earlier in the sample.\footnote{Such cases also provide insight into how \Providence is a sharper test in expectation because SO \BRAVO ignores information from the rest of the sample after the \BRAVO condition is met at some point earlier in the selection order.} EoR \BRAVO, however, does not stop. It might be difficult to explain to the public why SO \BRAVO stops in more extreme cases like this, where the condition is met early in the sample, but poor evidence for the alternative hypothesis in the rest of the sample is ignored. Second, only $5$ of the first $11$ ballots were for the announced winner. A first round of size $11$ would have resulted in a smaller average total number of ballots drawn, but would have provided a misleading sample (suggesting that the winner was incorrectly reported) due to a too-small sample size. 
Both these ideas are addressed more thoroughly in section~\ref{sec:workload}.
%first draft, needed to be shorter: ...stopping condition for the cumulative tally at the end of the round would see that the \BRAVO condition is not met (and would be correct that an EoR \BRAVO audit would require a second round for this sample). On the other hand, an SO \BRAVO audit does stop already in the first round because the cumulative sample met the \BRAVO stopping condition at some point before all the ballots had been added to the sample. In Figure~\ref{fig:pilot_sequence}, this is seen when the observed sample (orange) line touches (or goes above) the minimum winner ballots (blue) line which gives the \BRAVO stopping condition (i.e. a \BRAVO condition is met whenever the number of winner ballots in the sample is at least the number indicated by that blue line). While relatively rare, there also possible cases of selection order where the SO \BRAVO condition is met very early but then bad luck is had thereafter and the ultimate cumulative sample at the end of the round gives very poor evidence that the announced tally is correct. There could even be fewer ballots for the winner than the loser. Such cases might be difficult for an election official to explain to the public, especially when the evidence at the end of the round provides poor support for the alternative hypothesis.\footnote{Such cases also provide insight into how \Providence is a tighter test in expectation because SO \BRAVO ignores information from the rest of the sample after the \BRAVO condition is met at some point earlier in the selection order.} In Section~\ref{sec:workload} we address such cases more thoroughly.

\begin{figure}
\includegraphics[width=.5\textwidth]{pilot_sequence.png}
\caption{For each sample size from $1$ to $140$, the intermediate cumulative sum of ballots for the announced winner is shown, and the corresponding behavior is that of SO \BRAVO.}
\label{fig:pilot_sequence}
\end{figure}

>>>>

A first round size of $140$ ballots with large probability of stopping ($0.95$) was selected.
Selection order was tracked for the sake of analysis.
As expected, the audit concluded in the first round. The \Providence risk measure was $0.0418$. This is the smallest risk for which the sample would have passed a \Providence RLA for the announced election---the p-value of the statistical test and the inverse of the \Providence ratio defined in section \ref{sec:prov}. Table~\ref{tab:pilot-risks} shows risk measures for the drawn sample using \Providence, \Minerva and \BRAVO (both EoR and SO), all similarly defined.

\begin{table}[h!]
\begin{center}
\begin{tabular}{ |c|c|c|c|c| } 
\hline
%\diagbox[dir=NW]{First \\Round \\Size}{RLA}
ballots& \rotatebox{45}{\Providence} & \rotatebox{45}{\Minerva} & \rotatebox{45}{SO \BRAVO} & \rotatebox{45}{EoR \BRAVO} \\
\hline
140 & \bf{0.0418} & \bf{0.0418} & \bf{0.0541} & 0.366 \\
\hline
\end{tabular}
\end{center}
\caption{Risk measures for the drawn first round of $140$ ballots in the RI pilot audit. Risks in bold meet the risk-limit ($10\%$) and thus correspond to audits that would stop.}
\label{tab:pilot-risks}
\end{table}

@@

For the sake of analysis, the selection order of the ballots sampled during the pilot was also recorded. Figure~\ref{fig:pilot_sequence} shows the cumulative tally of winner ballots after each new ballot in the selection order is added to the sample.

We observe two interesting phenomena in this particular sample's selection order. 
First, an SO \BRAVO audit of this sample stops because the \BRAVO condition is met when the sample (orange line) surpasses the minimum number of winner ballots (blue line) earlier in the sample\footnote{Such cases also provide insight into how \Providence is a sharper test in expectation because SO \BRAVO ignores information from the rest of the sample after the \BRAVO condition is met at some point earlier in the selection order.}. EoR \BRAVO, however, does not stop at sample size $140$, the number of ballots drawn during the pilot. It might be difficult to explain to the public why SO \BRAVO stops in more extreme cases like this, where the condition is met early in the sample, but poor evidence for the alternative hypothesis in the rest of the sample is ignored. Second, only $5$ of the first $11$ ballots were for the announced winner. A first round of size $11$ would have resulted in a smaller average total number of ballots drawn, but would have provided a misleading sample (suggesting that the winner was incorrectly reported) due to a too-small sample size. 
Both these ideas are addressed more thoroughly in section~\ref{sec:workload}.
%first draft, needed to be shorter: ...stopping condition for the cumulative tally at the end of the round would see that the \BRAVO condition is not met (and would be correct that an EoR \BRAVO audit would require a second round for this sample). On the other hand, an SO \BRAVO audit does stop already in the first round because the cumulative sample met the \BRAVO stopping condition at some point before all the ballots had been added to the sample. In Figure~\ref{fig:pilot_sequence}, this is seen when the observed sample (orange) line touches (or goes above) the minimum winner ballots (blue) line which gives the \BRAVO stopping condition (i.e. a \BRAVO condition is met whenever the number of winner ballots in the sample is at least the number indicated by that blue line). While relatively rare, there also possible cases of selection order where the SO \BRAVO condition is met very early but then bad luck is had thereafter and the ultimate cumulative sample at the end of the round gives very poor evidence that the announced tally is correct. There could even be fewer ballots for the winner than the loser. Such cases might be difficult for an election official to explain to the public, especially when the evidence at the end of the round provides poor support for the alternative hypothesis.\footnote{Such cases also provide insight into how \Providence is a tighter test in expectation because SO \BRAVO ignores information from the rest of the sample after the \BRAVO condition is met at some point earlier in the selection order.} In Section~\ref{sec:workload} we address such cases more thoroughly.

\begin{figure}
\includegraphics[width=.5\textwidth]{pilot_sequence.png}
\caption{For each sample size from $1$ to $140$, the intermediate cumulative sum of ballots for the announced winner is shown. Observe that SO \BRAVO would stop because the \BRAVO condition (blue line) is achieved early on (orange line) in the audit. However, because the \BRAVO condition is no longer satisfied at $140$ ballots, an EoR \BRAVO audit would not have stopped. }
\label{fig:pilot_sequence}
\end{figure}
